The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-09 presentation/placenta_viviparous_presentation.Rnw
make figures larger and more centered

\begin{frame}{Species Studied}
% \begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\end{figure}
\end{frame}

\begin{frame}{Variation in Gestational Periods and Litter Size}
\footnotesize
%\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
%  \caption{Gestational periods in days and average liter sizes for studied species.}
%  \label{tab:gestational_periods}
%\end{table}
\end{frame}

\begin{frame}{Variations in Placenta Morphology}
%\begin{table}
\footnotesize
\begin{tabular}{c c c}
  \toprule
  Species & Interhemal Membrane & Pattern \\
  \midrule
  \smonkey & hemochorial & trabecular \\
  \cow     & epitheliochorial & villous \\
  \dog     & endotheliochorial & labyrinthine \\
  \armadillo & hemochorial & trabecular \\
  \horse & epitheliochorial & villous \\
  \elephant & endotheliochorial & labyrinthine \\
  \human & hemochorial & villous \\
% http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
  \opossum & epitheliochorial & choriovitelline \\
  \mouse & hemotrichorial & labyrinthine \\
  \galili & hemotrichorial & labyrinthine \\
  \sheep & epitheliochorial & villous \\
  \bonobo & hemochorial & villous \\
% http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
  \carmeli & hemotrichorial & labyrinthine \\
  \pig & epitheliochorial & diffuse \\
  \bottomrule
\end{tabular}
%\end{table}
\end{frame}

@@

\begin{frame}{Transcriptome Mapping Statistics: Reads}
\begin{figure}
<<transcriptome.mapping.figure,fig.width=10,fig.height=8,echo=FALSE>>=
print(ggplot(star.logs[grepl("input reads$",field),list(reads=as.numeric(value),species=capfirst(species))],
             aes(x=species,y=reads))
      + coord_flip()
      + geom_boxplot()
      + scale_y_log10(breaks=c(1e5,1e6,1e7,1e8,1e9,1e10),labels=trans_format("log10", math_format(10^.x)))
      + ylab("Reads per Sample")
      + xlab("Species")
      + theme_bw()
      + theme(legend.position="top",
              axis.text.x = element_text(size=10,face="italic"))
      )
@ 
\end{figure}
\end{frame}

\begin{frame}{Transcriptome Mapping Statistics: Unmapped}
\begin{figure}
<<transcriptome.mapping.figure.unmapped,fig.width=8,fig.height=8,echo=FALSE>>=
star.logs.unmapped <- 
    star.logs[grepl("%",field) & 
                  grepl("unmapped",field) & 
                      !grepl("too many mismatches",field),]
star.logs.unmapped <- 
    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
print(ggplot(star.logs.unmapped,
             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
      + geom_boxplot()
      + facet_grid(field~.)
      + ylab("Percent of Reads")
      + xlab("Species")
      + theme_bw()
      + coord_flip()
      + theme(legend.position="top",
              axis.text.x = element_text(face="italic"))
      )
@ 
\end{figure}
\end{frame}

@@

\begin{frame}{Core Placenta Transcriptome}
\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(),
              axis.text.y = element_text(face="italic"),
              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
      )
@ 
\end{figure}
\end{frame}

@@

\begin{frame}{Galectin Expression}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Galectins are major regulators of pregnancy in fetal and
  maternal tissues \cite{Than.ea2012:Galectinsguardiansofeutherian}
\item \textit{LGALS1} expressed everywhere
\item Primate expansion of \textit{LGALS13} and \textit{LGALS14}
\item Likely present in ancestral placenta of therian mammals as it is
  expressed even in \opossum
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

>>>>

\begin{frame}{Species Studied}
% \begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
% \end{figure}
\end{frame}

\begin{frame}{Variation in Gestational Periods and Litter Size}
\centering
\footnotesize
%\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
%  \caption{Gestational periods in days and average liter sizes for studied species.}
%  \label{tab:gestational_periods}
%\end{table}
\end{frame}

\begin{frame}{Variations in Placenta Morphology}
%\begin{table}
\centering
\footnotesize
\begin{tabular}{c c c}
  \toprule
  Species & Interhemal Membrane & Pattern \\
  \midrule
  \smonkey & hemochorial & trabecular \\
  \cow     & epitheliochorial & villous \\
  \dog     & endotheliochorial & labyrinthine \\
  \armadillo & hemochorial & trabecular \\
  \horse & epitheliochorial & villous \\
  \elephant & endotheliochorial & labyrinthine \\
  \human & hemochorial & villous \\
% http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
  \opossum & epitheliochorial & choriovitelline \\
  \mouse & hemotrichorial & labyrinthine \\
  \galili & hemotrichorial & labyrinthine \\
  \sheep & epitheliochorial & villous \\
  \bonobo & hemochorial & villous \\
% http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
  \carmeli & hemotrichorial & labyrinthine \\
  \pig & epitheliochorial & diffuse \\
  \bottomrule
\end{tabular}
%\end{table}
\end{frame}

@@

\begin{frame}{Transcriptome Mapping Statistics: Reads}
\begin{figure}
<<transcriptome.mapping.figure,fig.width=4,fig.height=4,echo=FALSE>>=
print(ggplot(star.logs[grepl("input reads$",field),list(reads=as.numeric(value),species=capfirst(species))],
             aes(x=species,y=reads))
      + coord_flip()
      + geom_boxplot()
      + scale_y_log10(breaks=c(1e5,1e6,1e7,1e8,1e9,1e10),labels=trans_format("log10", math_format(10^.x)))
      + ylab("Reads per Sample")
      + xlab("Species")
      + theme_bw()
      + theme(legend.position="top",
              axis.text.x = element_text(size=10),
              axis.text.y = element_text(size=10,face="italic"))
      )
@ 
\end{figure}
\end{frame}

\begin{frame}{Transcriptome Mapping Statistics: Unmapped}
\begin{figure}
<<transcriptome.mapping.figure.unmapped,fig.width=4,fig.height=4,echo=FALSE>>=
star.logs.unmapped <- 
    star.logs[grepl("%",field) & 
                  grepl("unmapped",field) & 
                      !grepl("too many mismatches",field),]
star.logs.unmapped <- 
    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
print(ggplot(star.logs.unmapped,
             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
      + geom_boxplot()
      + facet_grid(field~.)
      + ylab("Percent of Reads")
      + xlab("Species")
      + theme_bw()
      + coord_flip()
      + theme(legend.position="top",
              axis.text.x = element_text(size=10),
              axis.text.y = element_text(face="italic",size=10))
      )
@ 
\end{figure}
\end{frame}

@@

\begin{frame}{Core Placenta Transcriptome}
\begin{figure}
<<core.placenta.transcriptome,fig.width=10,fig.height=6,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
             aes(x=human_name,y=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + xlab("Human Ortholog Name")
      + ylab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(),
              axis.text.y = element_text(face="italic"),
              axis.text.x = element_text(angle=90,hjust=1,face="italic"))
      )
@ 
\end{figure}
\end{frame}

@@

\begin{frame}{Galectin Expression}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Galectins are major regulators of pregnancy in fetal and
  maternal tissues \cite{Than.ea2012:Galectinsguardiansofeutherian}
\item \textit{LGALS1} expressed everywhere
\item Primate expansion of \textit{LGALS13} and \textit{LGALS14}
\item Likely present in ancestral placenta of therian mammals as it is
  expressed even in \opossum
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<hormone.genes.plot,echo=FALSE,warning=FALSE,message=FALSE>>=
